Growing a Static Type System Inside a Python

### 2016-05-19 typy-2016/omar-026.tex
added initial abstract + intro + outline + dump for typy paper

%\documentclass[12pt]{article}
\documentclass[9pt,preprint]{sigplanconf}
\newcommand{\sigmat}[1]{\sigma_{\text{#1}}}

@@

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

\title{Extensible Gradual Typing Inside a Python}
%\subtitle{Unambiguously Growing a Static Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the static type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests the need for a mechanism that allows abstraction providers to extend the type system of a widely adopted language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.

Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library within Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax. They determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even when additional extensions are imported, so we avoid syntactic and semantic ambiguity by construction. We validate our approach with several examples of interesting type system fragments that are built into other languages but can be implemented as libraries using our approach, including functional constructs, a type system for secure string sanitation and a typed foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a library is easier than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several distinct languages often requires programming against their low-level implementation details, rather than their semantics, via a foreign-function interface (FFI) to a common implementation language (e.g. C or the JVM). This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries. As more languages are used together, more FFIs need to be constructed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues prevent even those dialects that introduce only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct and granular control over a language's syntax and semantics from within, or alongside, libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

@@

Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing and widely-used language, leaving its concrete and abstract syntax fixed. This choice is freeing: questions of syntactic ambiguity no longer need to be addressed. Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). These tools will continue to work. %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Our key insight is that few tools operate by exhaustive case analysis over type constructors. We introduce a mechanism that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete and abstract syntax, and the base semantics of the language, fixed.

To support new operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach). Instead, we delegate to a statically defined function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with these static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label), which maps to the abstract term $\FF{attr}[\text{``r''}](e)$ (i.e. the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$). A fragment defining statically typed pairs might see this term as referring to projection of the right component of the pair $e$, while a fragment defining functional record types might want it project out a field labeled \verb|r| from $e$, and an object system might introduce a more complex semantics involving dynamic dispatch. The type of dynamically classified Python values would follow Python's own complex field lookup protocol. Were we to naively ``open up'' the syntax and semantics, these fragments would conflict.

Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation. Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.

As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against (following the approach for composable syntax extensions as supported in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules''). %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

@@

\paragraph{Specific Aims} Our specific aims in this paper are to introduce active type constructors and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded. Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms. % We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e. normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience). 

@@

% We recommend abbrvnat bibliography style.
\newpage
\bibliographystyle{abbrv}

>>>>

%\documentclass[12pt]{article}
\documentclass[9pt]{sigplanconf}
\newcommand{\sigmat}[1]{\sigma_{\text{#1}}}

@@

\conferenceinfo{MODULARITY'15,}{March 16--19, 2015, Fort Collins, CO, USA}
\CopyrightYear{2015}
\crdata{978-1-4503-3249-1/15/03}

\title{Growing a Static Type System Inside a Python}
%\subtitle{Unambiguously Growing a Static Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using constructs available as libraries for the programming language they already use over those requiring a switch to a new language. But library providers have not been able to modularly extend the type systems of these languages, so some of the most powerful constructs have remained obscure. %Our aim in this paper is to introduce a mechanism that allows abstraction providers to extend the type system of a language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts. 
Our aims in this paper are to 1) introduce a mechanism that allows library providers to define new type system fragments modularly; 2) show how it can itself be implemented as a library, \texttt{typy}, for a widely used language, Python; 3) validate our approach with examples of non-trivial type system fragments that are built into other languages but that can be implemented as libraries using this mechanism, including functional constructs, an object system, a type system for secure string sanitation and a typed foreign function interface to OpenCL; and 4) formally specify the mechanism with a minimal typed lambda calculus, @$\lambda$. The key idea is to associate semantic logic with type constructors, rather than syntactic forms. 
%Both \texttt{typy} and @$\lambda$ have a fixed syntax. 
%The semantics of each term is determined by delegating control, during the process of typechecking and translating the term to the underlying \emph{internal language}, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even as additional type constructors are imported, so ambiguities cannot arise. 
% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

\category{D.3.2}{Programming Lan\-guages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
\keywords
type systems; extensible languages; metaprogramming
\section{Introduction}\label{intro}
Asking programmers to adopt a library is easier than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Yet researchers and domain experts continue to design new languages and language dialects, justifying this practice by  arguing that it would be  impossible to isomorphically express the constructs they need as libraries for existing languages. This is certainly a convincing argument when the constructs in question specify a non-trivial static semantics. %This practice is encouraged by the increasing availability of tools like language workbenches \cite{erdweg2013state}. 
Unfortunately, the flexibility afforded taking such a \emph{language-oriented approach} \cite{journals/stp/Ward94} to new ideas comes at the cost of modularity: working with a component written in a different language typically requires programming against the low-level details of a particular implementation of that language, rather than its semantics, via a foreign-function interface (FFI) (e.g. to C, or JVM bytecode). This is brittle (the implementation may change), unnatural and unsafe. Mitigating the loss of static safety by adding run-time checks at language boundaries has a performance cost. %As more languages are used together, more FFIs need to be constructed and traversed.% This can defeat the productivity gains afforded by these dialects.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These problems are relevant even to dialects that introduce only a few new constructs. A study comparing a Java dialect, Habanero-Java (HJ), with a  library, \lstinlinep{java.util.concurrent}, found that the language-based abstractions in HJ were indeed easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This appears to be a widespread problem: programmers and development teams often cannot use  the constructs they might prefer because they are only available in specialized dialects they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Language extension mechanisms have long promised to reduce these problems by  giving providers the ability to define syntactic and semantic constructs more granularly, so that clients can import them as needed. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often housed in an obscure language dialect and so face the classic ``chicken-and-egg'' problem: languages like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit syntax extensions over a fixed semantics) must overcome the same sorts of  barriers to adoption as a language like HJ. Second, they give providers too much control over the language, again at the cost of modularity, because providers can define conflicting constructs. The conflicts, if detected at all, are detected too late: when a client attempts to use the conflicting constructs in the same program. %Clients are not generally capable of understanding and fixing the issue productively.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

@@

In this paper, we take steps toward addressing both of these problems by introducing a mechanism that allows library providers to define new type system fragments modularly, and showing how it can be itself be implemented as a library, \texttt{typy}, for an existing and widely-used language, Python (using its quasiquotation and reflection facilities). This constrains us: we do not have the power to change Python's concrete and abstract syntax. But in fact, we find that this constraint  can be freeing: questions of syntactic conflict never arise, and tools that operate by exhaustive case analysis over the syntax (e.g. syntax highlighters, editor plugins, pretty-printers, style checkers, documentation generators) continue to work. %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

Library providers introduce type system fragments by defining new type constructors (\emph{tycons}, for brevity). For example, a library provider might introduce record types, as are built in to functional languages like ML and Haskell, by defining a new tycon, \lstinline{record}. A different library provider may introduce a statically-typed variant of an object system featuring prototypic inheritance (al\'a Javascript) by defining the tycon \lstinline{proto}. Types are constructed by applying a tycon to a statically valued \emph{type index}, written \lstinline{tycon[tyidx]}. For example, we can construct a record type with two rows like this:
\begin{lstlisting}[numbers=none]
record[{
  'label1' : dyn[()],    
  'label2' : string[()]
}]
\end{lstlisting}
The index is a dictionary mapping row labels to their corresponding types. Here, \lstinline{dyn} (which we discuss later) and \lstinline{string} are trivially indexed tycons, i.e. the only valid index is \lstinline{()}. For concision, trivial indices can simply be omitted (we showed them here only to emphasize the uniform manner in which one should think about type construction). A prototypic object type having an analagous field structure (and for simplicity, assuming the default prototype) is constructed like this:
\begin{lstlisting}[numbers=none]
proto[{ # (fields of a prototypic object are mutable)
  'label1': dyn,   # equivalent to dyn[()]
  'label2': string # equivalent to string[()]
}]
\end{lstlisting}

Given that a programmer might be using both of these libraries in the same program, how should our system determine the semantics of an expression like \lstinline{e.label1}, where \lstinline{e} is some sub-expression? In a monolithic language, the usual approach is to define the semantics in a \emph{syntax-directed} manner: the language specifies, in one place, all the logic relevant to each form. Our approach, in contrast, is \emph{tycon-directed}: the language delegates  to a \emph{statically valued function} defined by some relevant tycon, called the \emph{delegate}. The language specifies only the \emph{delegation protocol} that assigns a delegate to each form. In this example, the delegate is the tycon of the type recursively synthesized by \lstinline{e}. We will see how tycons like \lstinline{record} and \lstinline{proto} statically compute a type and a translation (to Python code) for such a term when they are delegated control by this protocol in Sec. \ref{usage}. For now, it suffices to point out that the code involved is almost exactly what one would write if implementing a language that built in record types directly.

% A statically-valued function 
% %, identified based on the name Python's grammar gives to the form, here \lstinline{syn_Attribute},  %So if \lstinline{e} is of record type, then the \lstinline{record} tycon  is delegated control, and if \lstinline{e} is of prototypic object type then \lstinline{proto} is delegated control, in either case by calling \lstinline{syn_Attribute} with a reference to the syntax tree of the term as well as a \emph{typing context}, which we will describe later. 
% must  either synthesize a type for the term as a whole or indicate a type error. If it synthesizes a type, the tycon must then generate a \emph{translation} for the term as a whole, which determines its dynamic behavior. So, for the term \lstinline{e.label1}, if \lstinline{e} is of record type, then the \lstinline{record} tycon is delegated control. It will check that \lstinline{label1} is one of the row labels specified in the index of the type of \lstinline{e}. Records can be implemented by translation to tuples (the row labels are only needed statically), so the translation it then generates will project out the appropriate position from the translation recursively generated for \lstinline{e}.

% We will detail how tycons are implemented as Python classes in Sec. \ref{usage}.  The innovation is in how this protocol decentralizes control over a fixed set of forms, allowing us to define a range of semantic constructs in libraries. The language itself needs to specify only the \emph{delegation protocol} that determines how a tycon is delegated control over each syntactic form.

% More specifically, what this means is that the statically valued function  because that the tycon defines must either synthesize a type for the term as a whole or indicate a type error. If a type is synthesized, the tycon must also generate a \emph{translation} for the term as a whole.

% We call a type constructor equipped with a set of such statically valued functions an \emph{active type constructor}.

% More specifically, the semantics are a \emph{bidirectionally-typed type-directed translation semantics}.

% allow the associated operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach). Instead, we delegate to a statically defined function associated with a relevant type constructor. The term constructor's identity only serves to determine how to determine this delegate. For example, consider the concrete term \lstinlinep{x.r} (where \lstinlinep{x} is a variable and \lstinlinep{r} is a static label), which, according to Python's grammar, maps to the abstract term \lstinlinep{Attribute("r"; Name("x"))} (i.e. the term constructor $\FF{Attribute}$, \emph{indexed} by a static representation of the label and one \emph{argument}, $e$). There are many possible meanings for this term. If $e$ is a pair, this might be a way to project out its right component. If $e$ were a functional record,  this might be the way to project out a field labeled \verb|r|. An object system might introduce a more complex semantics involving dynamic dispatch. The type of dynamically classified Python values would follow Python's own complex field lookup protocol. Were we to naively ``open up'' the syntax and semantics in an attempt to allow all of these features into our language, these semantics would conflict.

% Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation. Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.

This example was of a form for which there is a clear ``target'' subexpression that can be examined recursively to determine the tycon to delegate control to. For other forms, it is perhaps less clear. For example, the curly-brace-delimited form in the following expression could be given a record type, a prototypic object type or be a dynamically tagged Python dictionary, in which case it would have type \lstinline{dyn}:
\begin{lstlisting}[numbers=none]
f({label1: "val1", label2: "val2"})
\end{lstlisting}
We resolve this ambiguity by leveraging the \emph{local type inference} mechanism that \verb|typy| uses \cite{Pierce:2000:LTI:345099.345100}. That is, for literal forms in Python's grammar, the tycon of the type that the term is being analyzed against is delegated control. Here, if \lstinline{f} is a function taking an argument of the record type defined above, then \lstinline{record} is delegated control over the curly-brace-delimited form. This can be understood as a variant of the mechanism that Wyvern introduced for resolving ambiguous syntax extensions \cite{TSLs}, here applied in a setting where there is a fixed syntax but an extensible semantics.% When a literal form appears at the top-level of a function, where there is no expected type, the programmer can provide a type annotation on an expression directly, using the form \lstinline{e [: T]}, where \lstinline{e} is an expression and \lstinline{T} is a type (or a tycon applied to an incomplete index, which we discuss in Sec. \todo{inc indices}).

Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss.

There is always a single tycon delegated  responsibility over a term and this delegate is stable with respect to additional library imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules''). %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

@@

\paragraph{Specific Aims} Our specific aims in this paper are to introduce this approach, which we call \emph{active typechecking and translation}, and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded. Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms. % We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e. normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience). 

@@

% We recommend abbrvnat bibliography style.
%\newpage
\bibliographystyle{abbrv}
